Fusing Filters with Integer Linear Programming

### 2014-05-08 2014betterfusionforfilters/section/03-SizeInference.tex
Add Nat types to dep type version of signature
Add note for clock inference
Minor stuff

%!TEX root = ../Main.tex
\section{Size inference}
Before performing fusion proper, we must infer the relative sizes of each array in the program. We achieve this with a simple constraint based inference algorithm, which we discuss in this section. Size inference has been previously described in the context of array fusion by Chatterjee~\cite{chatterjee1991size}, but this system does not support size changing functions such as filter.

@@

When giving the size scheme of an operator graph we use $:_s$ to indicate that this is a size scheme rather than a type scheme.

@@

Finally, note that size schemes form but one ``layer'' of the type information that would be expressible in a full dependently typed language. For example, in Coq or Agda~\CITE we could write something like:
\begin{alltt}
filterLeft 
 : \(\forall\,k\sb{1}.\,\exists\,k\sb{2}.\) Array \(k\sb{1}\) Float
              \(\to\) (Array \(k\sb{1}\) Float, Array \(k\sb{2}\) Float)
\end{alltt}

@@

% -----------------------------------------------------------------------------
\subsection{Constraint Generation}
The rules for constraint generation are shown in Figure~\ref{f:ConstraintGeneration}. The judgment ~~$\SizeB{\Gamma_1}{zs}{b}{\Gamma_2}{C}$ reads: ``Under environment~$\Gamma_1$, array variable $zs$ binds the result of $b$, producing a new environment $\Gamma_2$ and size constraints $C$''. The other judgment forms are similar. The environment $\Gamma$ has the following grammar:
$$
\Gamma @::=@ ~~\cdot ~~|~~ \Gamma,~ \Gamma ~~|~~ zs : k ~~|~~ k ~~|~~ \exists k
$$

The $\cdot$ and $\Gamma,~ \Gamma$ are the empty environment and environment concatenation as usual. The element $zs : k$ records the size $k$ of some array variable $zs$. A plain $k$ indicates that $k$ can be unified with other size types when solving constraints, whereas $\exists k$ indicates a ``rigid'' size variable that cannot be unified with other sizes. We use the $\exists k$ syntax because this variable will also be existentially quantified if it appears in the size scheme of the overall function.

@@

For example, the final environment and constraints generated for the @normalize2@ example from Section~\ref{s:Introduction} are as follows:
$$
\begin{array}{ll}
       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2 
\\
\vdash & true 
        ~\wedge~  k_{gts} = k_1
        ~\wedge~  true
\\     &~~~~~~~~ 
          \wedge~  k_{xs}  ~= k_2
        ~ \wedge~  k_{ys1}  = k_2 
        ~ \wedge~  k_{xs}   = k_3
        ~ \wedge~  k_{ys2}  = k_3
\end{array}
$$

@@

>>>>

%!TEX root = ../Main.tex
\section{Size inference}
Before performing fusion proper, we must infer the relative sizes of each array in the program. We achieve this with a simple constraint based inference algorithm, which we discuss in this section. Size inference has been previously described in the context of array fusion by Chatterjee~\cite{chatterjee1991size}. In constrast to our algorithm, \cite{chatterjee1991size} does not support size changing functions such as filter.

@@

We write $:_s$ to distinguish size schemata from type schemata.

@@

Finally, note that size schemes form but one ``layer'' of the type information that would be expressible in a full dependently typed language. For example, in Coq or Agda~\CITE we could write something like:
\begin{alltt}
filterLeft : \(\forall\,k\sb{1}:\,\)Nat\(.\,\exists\,k\sb{2}:\,\)Nat\(.\) 
  Array \(k\sb{1}\) Float \(\to\) (Array \(k\sb{1}\) Float, Array \(k\sb{2}\) Float)
\end{alltt}

@@

% -----------------------------------------------------------------------------
\subsection{Constraint Generation}
The rules for constraint generation are shown in Figure~\ref{f:ConstraintGeneration}. The judgment ~~$\SizeB{\Gamma_1}{zs}{b}{\Gamma_2}{C}$ reads: ``Under environment~$\Gamma_1$, array variable $zs$ binds the result of $b$, producing a new environment $\Gamma_2$ and size constraints $C$''. The other judgment forms are similar. The environment $\Gamma$ has the following grammar:
$$
\Gamma~ @::=@ ~~\cdot ~~|~~ \Gamma,~ \Gamma ~~|~~ zs : k ~~|~~ k ~~|~~ \exists k
$$

As usual, $\cdot$ represents the empty environment and  $\Gamma,~ \Gamma$ environment concatenation. The element $zs : k$ records the size $k$ of some array variable $zs$. A plain $k$ indicates that $k$ can be unified with other size types when solving constraints, whereas $\exists k$ indicates a ``rigid'' size variable that cannot be unified with other sizes. We use the $\exists k$ syntax because this variable will also be existentially quantified if it appears in the size scheme of the overall function.

@@

For example, the final environment and constraints generated for the @normalize2@ example from Section~\ref{s:Introduction} are as follows:
$$
\begin{array}{ll}
       & x : k_{xs},~ gts : k_{gts},~ \exists k_1,~ k_2,~ k_3 
\\
\vdash & true 
        ~\wedge~  k_{gts} = k_1
        ~\wedge~  true
\\     &~~~~~~~~ 
          \wedge~  k_{xs}  ~= k_2
        ~ \wedge~  k_{ys1}  = k_2 
        ~ \wedge~  k_{xs}   = k_3
        ~ \wedge~  k_{ys2}  = k_3
\end{array}
$$

@@

\TODO{Cite work on clock inference for data flow languages. Eg http://www.irisa.fr/prive/talpin/papers/hldvt05.pdf}
